---
title: "Getting started with clinstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clinstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clinstats)
```

`clinstats` collects the statistical routines that recur in clinical research
workflows: discrimination summaries, odds ratios, regression tables, survival
cut points, and expression-matrix preparation. Every function returns a
tibble, so results can be printed for a report or filtered, joined, and
exported programmatically.

## Bundled data

The package ships a curated colorectal cancer clinical dataset assembled from
public GEO series and TCGA exports.

```{r}
dim(clin_crc)
table(clin_crc$dataset)
```

The three stored versions (`clin_crc`, `clin_crc_gf`, `clin_crc_cell`)
correspond to the 2018 and 2019 harmonisations of the original repository.
See the dataset help page for the column layout.

## Regression tables

`logistic_table()` and `cox_table()` fit every variable on its own, fit one
multivariable model, and return a single table with both sets of estimates.

```{r}
logistic_table(
  clin_crc,
  outcome = "rfs.event",
  factors = c("sex", "age", "tnm.stage", "cms")
)
```

```{r}
cox_table(
  clin_crc,
  time = "rfs.delay",
  event = "rfs.event",
  factors = c("sex", "age", "tnm.stage", "cms")
)
```

The multivariable model includes all requested variables by default. Pass
`multivariable = "significant"` to keep the historical screening rule in
which only variables with a univariable likelihood-ratio p-value below 0.05
enter the model.

## Marker evaluation

`roc_summary()` reports the area under the curve together with the optimal
operating point, and `odds_ratio()` reports the odds ratio for the high group
at the Youden-optimal threshold.

```{r}
marker <- scale(clin_crc$age)
roc_summary(marker, clin_crc$rfs.event)
odds_ratio(marker, clin_crc$rfs.event)
```

For a survival endpoint, `survival_cutoff()` estimates the Kaplan-Meier
Youden index over a grid of candidate cut points and returns the best one.

```{r}
survival_cutoff(
  time = clin_crc$rfs.delay,
  event = clin_crc$rfs.event,
  marker = marker,
  predict_time = 3
)
```

## Expression data

`clean_expression()` prepares an expression matrix: it drops unusable
columns, expands `" /// "` multi-mappings, optionally maps identifiers with
`org.Hs.eg.db`, and collapses duplicated gene columns to the probe with the
largest median absolute deviation.

```{r}
expr <- matrix(
  rnorm(6 * 4),
  nrow = 6,
  dimnames = list(NULL, c("A", "A", "B", "C /// D"))
)
clean_expression(expr)
```

`oncotype_crc()` computes the Oncotype DX colon cancer recurrence score from
the published panel; genes are expected in rows.

## Reproducibility

The bundled datasets are byte-identical to the archived
`curatedClinicalData` objects, with provenance and checksums recorded in
`data-raw/README.md`. Statistical conclusions drawn from them remain the
responsibility of the analyst.
